---
title: "Reproducible Research project 1"
author: "Roderic Reckley"
date: "February 7, 2016"
output: html_document
---


## Loading and preprocessing the data
The first step is to include the necessary libraries and idenitfy the location of the data file to be read in.
The data is read in via the read.csv function.
```{r}
library(ggplot2)
library(lattice)
setwd("C:/Users/RNRECKLEY/Downloads/coursera/datascience/reproducible/project1")
activity_data <- read.csv("activity.csv")
activity_data$date <- as.Date(activity_data$date)
```


```{r, echo=FALSE}
print(paste("The total number of observations found in the activity data set is : ", nrow(activity_data)))
```



## What is mean total number of steps taken per day?
The first step is to remove the observations with the missing values from the data set.
Once this is done generate the histogram using the ggplot function.

```{r}
cleaned_data <- na.omit(activity_data)
```

```{r, echo=FALSE}
print(paste("The number of observations left in the  activity data set after removing those with missing values is : ", nrow(cleaned_data)))
```



```{r, echo=FALSE}
plot1 <- ggplot(cleaned_data, aes(date, steps)) 
plot1 <- plot1 + geom_bar(stat = "identity") +
     labs(title = "Histogram of Total Number of Steps Taken Each Day\n with missing values removed", x = "Date", y = "Total number of steps")
print(plot1)
```

```{r}
sum_data <- aggregate(activity_data$steps, by=list(activity_data$date), FUN=sum, na.rm=TRUE)
names(sum_data) <- c("Date", "TotalSteps")
```


#### Find the mean and medial values of the activity data set.
```{r, echo=FALSE}
print(paste("The mean number of steps taken per day with missing data is ", round(mean(sum_data$TotalSteps))))

print(paste("The median number of steps taken per with the missing data is ", round(median(sum_data$TotalSteps))))
```


```{r}
cleaned_sum_data <- aggregate(cleaned_data$steps, by=list(cleaned_data$date), FUN=sum, na.rm=TRUE)
names(cleaned_sum_data) <- c("Date", "TotalSteps")
```

```{r, echo=FALSE}
print(paste("The mean number of steps taken per day without the missing data is ", round(mean(cleaned_sum_data$TotalSteps))))

print(paste("The median number of steps taken per without the missing data is ", round(median(cleaned_sum_data$TotalSteps))))
```




## What is the average daily activity pattern?
```{r}
avg_daily <- aggregate(cleaned_data$steps, list(interval = as.numeric(as.character(cleaned_data$interval))), FUN = "mean")
names(avg_daily)[2] <- "AverageSteps"

```


```{r, echo=FALSE}
plot2 <- ggplot(avg_daily, aes(x=interval, y=AverageSteps)) +   
  geom_line(color="red", size=1) +  
  labs(title="Average Daily Activity", x=" Time Interval (5-minutes)", y="Number of steps") +  
  theme_bw()
print(plot2)

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_interval <- avg_daily[which.max(avg_daily$AverageSteps),]
print(max_interval)
```

The maximum number of steps observered across all days occurs at interval 835 with an average of 206.1698 steps.

## Imputing missing values

```{r, echo=FALSE}
print(paste("The total number of missing values in the actvity data set is ", sum(is.na(activity_data))))
```

```{r}
sum(is.na(activity_data))
imputed_activity <- merge(activity_data, avg_daily, by="interval")
missing_values <- which(is.na(imputed_activity$steps))
imputed_activity$steps[missing_values] <- imputed_activity$AverageSteps[missing_values]
imputed_activity <- imputed_activity[,c(1:3)]
sum(is.na(imputed_activity))

```



```{r, echo=FALSE}
plot3 <- ggplot(imputed_activity, aes(date, steps)) 
plot3 <- plot3 + geom_bar(stat = "identity") +
  labs(title = "Histogram of Total Number of Steps Taken Each Day\n with missing value replaced by mean
      ", x = "Date", y = "Total number of steps")
print(plot3)

```

```{r}
imputed_steps_taken <- aggregate(imputed_activity$steps, by=list(imputed_activity$date), FUN=sum)
names(imputed_steps_taken) <- c("Date", "TotalSteps")

```

#### Find the mean and medial values of the activity data after imputing the missing data.
```{r, echo=FALSE}
print(paste("The mean number of steps taken per day after imputing the missing data is ", round(mean(imputed_steps_taken$TotalSteps))))

print(paste("The median number of steps taken per day after imputing the missing data is ", round(median(imputed_steps_taken$TotalSteps))))

```
#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
  If the observations with missing data are included when calculating the mean and median, imputing the data has a significant impact on the calculation of the mean and meadian. In this instance, the mean increases from 
9354 to 10766. The median inceases from 10395 to 10766.

  If the observastions with missing data are exluded from calculating the mean and median, imputing the data has no impact. The mean and median generated approximately the same. 



## Are there differences in activity patterns between weekdays and weekends?
```{r}
imputed_activity$Day <- factor(format(imputed_activity$date, "%A"))
levels(imputed_activity$Day)
levels(imputed_activity$Day) <- list(weekday = c("Monday", "Tuesday","Wednesday","Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(imputed_activity$Day)
table(imputed_activity$Day)

steps_taken_by_day <- aggregate(imputed_activity$steps,
                                list(interval= as.numeric(as.character(imputed_activity$interval)),
                                     weekdays = imputed_activity$Day), FUN = "mean")

names(steps_taken_by_day)[3] <- "AvgSteps"


```



```{r, echo=FALSE}
plot4 <- xyplot(steps_taken_by_day$AvgSteps ~ steps_taken_by_day$interval | steps_taken_by_day$weekdays, 
                layout = c(1, 2), type = "l", 
                xlab = "Interval", ylab = "Average Number of steps")
print(plot4)
```

   When examing the graph of the weekend and weekday activtiy, several distinct patterns emerge. 
 First, the weekdays has the peak or largest average number of steps is observed. Here the peak occurs during 
 the morning and drops off significantly duing the middle of the day. Minor peaks of activity are observered during the early afternoon and evening hours of a weekday. With the weekend, the activity peak occurs during the same interval, but the average number of steps observered during this period is less. Unlike weekday's, there is more activity occured observered after the initial peak. The weekend activity appears to spread out during the majority of the day.


